---
title: "Introduction to talusR"
author:
  - name: "Chao-Jen Wong"
    affiliation: "Seattle Children's Research Institute"
    email: chao-jen.wong@seattlechildrens.org
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to talusR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  eval = TRUE, 
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = 'center',
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Getting started

For now, this package only takes Talus data that are pre-processed by DIANN—i.e. the summarized protein-abundance file (`*.tsv`). The input metadata table (`*.csv`) must include a column (`which_run`) whose entries exactly match the sample (run) names in that abundance file.

This package also assumes that Talus process only one sample in a single run (one mass-spectrometry acquisition), and DIANN has executed the normalization process across all the runs. 

Install `talusR`:
```{r install-talusR, eval=FALSE}
devtools::install_github('chaochaowong/talusR',
                         ref          = "dev",
                         dependencies = TRUE
)

# or to install the Bioconductor dependencies
BiocManager::install(
  "chaochaowong/talusR@dev",
  dependencies = TRUE
)
```

```{r load-library}
library(talusR)
```

# Import Talus data








