---
title: "Introduction to talusR"
author:
  - name: "Chao-Jen Wong"
    affiliation: "Sarthy Lab, Seattle Children's Research Institute"
    email: chao-jen.wong@seattlechildrens.org
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to talusR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  eval = TRUE, 
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = 'center',
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Getting started

## Intruduction
The `talusR` package provides a workflow for Talus mass-spectrum data—import, Qc, differential analysis, and visualization. But it assumes that your data have already been processed by DIA-NN (e.g. `*.tsv` or `*.csv` outputs containing normalized, protein-level intensity values). It further expects one sample per mass-spectrometer run (with DIA-NN normalization applied across all runs), and that your sample metadata table includes a run column whose entries exactly match the sample names in the intensity signal file. 

Key features include:

- **Data import & structure**  
  Load DIA-NN–processed protein intensities into S4 objects:  

  - `TalusDataSet` for a whole dataset  
  - `TalusDataSetList` (a `SimpleList` of `TalusDataSet`) for per-fraction splits  

  Each object carries sample metadata and feature annotations.

- __Quality control__: Run PCA or other multivariate checks to confirm that technical replicates and fractions cluster as expected.    
- __Differential analysis__: use a simple wrapper around `matrixTests` to t-test and `limma` to fit linear models, define contrasts, apply independent filtering by adjusted p-value or log-fold change thresholds, and return results in `TalusResult` or `TalusResultList` objects.
- __Visualization__: Create PCA scatterplots, volcano plots, per-protein intensity plots.
- __Bioconductor integration__: Built on `SummarizedExperiment` and `S4Vectors`, `talusR` classes inherit the generics and methods of the Bioconductor ecosystem, for convinient plug into annotation, pathway analysis, or reporting tools.


## Installation

```{r install-talusR, eval=FALSE}
devtools::install_github('chaochaowong/talusR',
                         ref          = "dev",
                         build_vignettes = TRUE)

# or install along with the Bioconductor dependencies with appropriate version
BiocManager::install("chaochaowong/talusR@dev",
                     dependencies = TRUE
                     build_vignettes = TRUE
)
```

## Loading libraries

```{r load-library}
library(talusR)
library(ggplot2)
library(dplyr)
```

# Import Talus data
The core function `read_talus()` imports protein-level intensity signals for whole dataset or, if `split_by_fraction = TRUE`, splits by fractions.

## Whole dataset
When `split_by_fraction = FALSE`, `read_talus()` returns a single `TalusDataSet` instance.  This mode is ideal when:

- Your data have no fraction annotations       
- You want to run PCA-based QC to verify that protein-level signals from each fraction cluster as expected        

```{r assemble-files}
# get paths
protein_file <- system.file('extdata', 'test_data.tsv', 
                            package = 'talusR')
meta_file    <- system.file('extdata', 'test_meta.csv',
                            package = 'talusR')
```
```{r import-test-data}
# without split by fraction -> return TalusDataSet
tds  <- talusR::read_talus(file = protein_file,
                           meta_file = meta_file,
                           which_proteinid = "Protein.Ids",
                           which_run = "Run",
                           remove_few_measurements = TRUE,
                           split_by_fraction = FALSE)
tds
```


## Per-fraction splits
When `split_by_fraction = TRUE`, `read_talus()` returns a `TalusDataSetList` instance—a `SimpleList` of `TalusDataSet` instances, one per fraction, each holding its protein-level signals.

```{r split-by-fraction}
tdsl <- read_talus(file = protein_file,
                   meta_file = meta_file,
                   which_proteinid = "Protein.Ids",
                   which_fraction = "Frx",
                   which_sequence = NA,
                   which_run = "Run",
                   remove_few_measurements = TRUE,
                   split_by_fraction = TRUE,
                   log_transform = 'log2',
                   intensity_group = "protein",
                   metric = "DIA-NN")

tdsl
```

# QC
Use `plot_pca()` to assess sample clustering and variance structure.

## `TalusDataSet`
Run PCA on the full dataset to verify that protein-level signals from each fraction cluster as expected.
```{r pca-tds, fig.cap='PCA on TalusDataSet.'}
talusR::plot_pca(tds, color_by='Frx')
```

## `TalusDataSetList`
When using per-fraction splits, perform PCA on each fraction independently:
```{r pca-for-TalusDataSetList, fig.cap='PCA performed separately on each fraction in the TalusDataSetList instance.'}
plot_pca(tdsl, color_by='Tx')
```

# Differential analysis and visualization

**Core statistical methods:**

- `talus_limma`: a wrapper function for `limma::lmFit()` to fit linear models and test contrasts.  
- `talus_row_t_welch`: a wrapper function for `matrixTests::row_t_welch()` to perform per-protein Welch’s t-tests.

**Summary & visualization:**

- `summary_talus`: summarize differential results, flag significant proteins, and return a `data.frame` of significant proteins.
- `plot_volcano`: create a volcano plot of log-fold changes vs adjusted p-values.  
- `plot_per_protein`: display protein intensity profiles across fractions or conditions.  


## `limma`
Use `talus_limma()` to fit linear models with empirical Bayesian and moderate t-statistics across conditions (e.g. drug treatment) to perform differential analysis. 

Here the condition group is `Tx`: _DMSO_, _etopo_, and _doxo_. Set the factor levels so that  _DMSO_ is the reference:
```{r define-tx-levels}
tx_levels <- c('DMSO', 'Etopo', 'Doxo')
tdsl <- endoapply(tdsl, function(x) {
  colData(x)$Tx <- factor(colData(x)$Tx, levels = tx_levels)
  x
})
```

Next, run `talus_limma()` with a `~ 0 + Tx` design formula to compare each treatment against the `DMSO` control:
```{r limma}
res_limma <- talus_limma(tdsl, design = ~ 0 + Tx)
```

Finally, summarize and filter for significant proteins with with adjusted p-value < 0.05 and |log₂ fold change| > 0.5.
```{r summarize-limma}
limma_filt <- summary_talus(res_limma[['chrom']], 
                            alpha = 0.05, 
                            lfc_threshold = 0.5)
limma_filt
```

Visualize the statistiics by `plot_volcano()`:
```{r volcano-limma}
plot_volcano(res_limma[['chrom']], 
             alpha = 0.05,
             lfc_threshold = 0.5, 
             use_adjP = TRUE, 
             label_top_n=10)
```

## t-test

Apply `talus_row_t_welch()` to perform per-protein Welch’s t-tests on the `TalusDataSetList`:

```{r row-t-welch}
res_t <- talus_row_t_welch(tdsl, design = ~ 0 + Tx)
names(res_t)
```

Filter significant proteins using summary_talus() (adjusted p-value < 0.05, |log₂FC| > 0.5):

```{r show-row-t-summary}
# show contrast level
cat('\nContrast levels:\n')
names(res_t[['chrom']])

# summarize the result per fraction
ttest_filt <- summary_talus(res_t[['chrom']], 
                            alpha = 0.05, 
                            lfc_threshold = 0.5)
```

Visualize the statistics by `plot_volcano()`. 
```{r viz-t-welch-by-volcano}
# try to re-arrange the contrast_levels
plot_volcano(res_t[['chrom']], 
             alpha = 0.05,
             lfc_threshold = 0.5, 
             use_adjP = TRUE,
             contrast_levels = names(res_t[['chrom']])[c(2, 1)],
             label_top_n=10)
```

## Per-protein plot
Extract a protein ID (e.g. TOP2A) and plot the intensities across conditions (e.g., fractions and treatments):
```{r plot-per-protein}
protein_id <- as.data.frame(rowData(tdsl[[1]])) %>% 
  dplyr::filter(Genes == 'TOP2A') %>% pull(Protein.Group)

plot_per_protein(tdsl, 
                 protein_id = 'P11388;P11388-2;P11388-3;P11388-4',
                 group_by = 'Tx') +
  labs(title = 'TOP2A')
```

Or just for _chrom_ fraction only:
```{r plot-per-protein-chrom}
plot_per_protein(tdsl[['chrom']], 
                 protein_id = 'P11388;P11388-2;P11388-3;P11388-4',
                 group_by = 'Tx') +
  labs(title = 'TOP2A')
```


# sessionInfo
```{r sessioninfo}
sessionInfo()
```


